---
title: "Final Project"
author: "Group1"
date: "2022-11-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load-packages, message=FALSE}
library(tidyverse)
library(knitr)
```

Variable Descriptions: 
- instant: record index 
- dteday: date 
- season: season (winter, spring, summer, fall) 
- yr: year (0: 2011, 1:2012) 
- mnth: month (1 to 12) 
- holiday: weather day is holiday or not 
- weekday: day of the week 
- workingday: if day is neither weekend nor holiday is 1, otherwise is 0. 
- weathersit: clear, mist, or rain or snow 
- temp: Normalized temperature in Celsius (range from 0 to 1) 
- atemp: Normalized feeling temperature in Celsius 
- hum: Normalized humidity.  
- windspeed: Normalized wind speed. The values are divided to 67 (max) 
- casual: count of casual (unregistered) users 
- registered: count of registered users 
- casual_percent: percent of daily bike users who are not registered users 
- cnt: count of total rental bikes including both casual and registered 


```{r load-data}
bike_data <- read.csv("bike_share.csv")
```

```{r}
#add your code here

ggplot(data=bike_data, aes(x = atemp, y = cnt), size=1) +
  geom_boxplot() +
  labs(x="Temperature", y="Users")+
  theme_bw()

```

```{r scatterplot, echo=FALSE}
ggplot(data=bike_data)+
  geom_point(aes(x=atemp,y=cnt),size=1)+
  labs(x="Feeling Temperature",y="Number of users")+
  theme_bw()
```

